|

### 2021-03-24 paper/v2_taylor_francis/broadband.speed.covid.Rmd
Most changes discussed complete.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds
during the spring 2020 lockdown,
may reinforce or redress the spatial and social dimensions of digital divisions.
We employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and interactions
between digital divisions geographically and socioeconomically.
Our analysis enables us to better understand how the spatial and social distribution
of education, occupation and online accessibility intersect to enable or hinder 
the practice of telecommuting at a time of extreme demand.
We also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era.

@@

Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
In some aspects, these returns may have increased during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies. Rural areas, predictably, fare worse.
This finding broadly applies in the UK, where studies that also analysed broadband 
speed checks concluded that average speeds are lower in rural areas, something that
has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
Moreover, @riddlesden2014broadband found that levels of deprivation did not correlate with 
first level digital divides.

However, as @blank2018local highlight, variation
in individual internet uptake and use is a product of more complex spatial and 
demographic characteristics than simple rurality or urbanisation.
Dense urban areas were shown to suffer more from slowdown during peak hours, 
although these services were more likely to be improved by increased competition 
between providers, such as between new entrants and Virgin Media cable connections
[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
available to only $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider. 
Whether the variation in infrastructure quality and reliability affects the 
spatial footprint of telecommuting has not previously been investigated, in part 
because telecommuting has not previously been a cause of concentrated demand and 
pressure on internet services. Other intersections between digital divides have been
subject to study.
There are indications that those who purchase high speed connections consume
more data of all sorts and use their connections for a greater variety of purposes
[@hauge2011consumer].
There is also a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
These studies suggest that better internet services enable households to make 
savings and efficiencies, an example of the first level digital divide reinforcing 
the third level.

<!-- spatio-temporal -->
Currently there is a hotly debated discussion in the literature regarding these 
exact changes in transportation and the structure of cities. We now have enough 
hard data to observe the drastic change in the space-time geography of cities around 
the world [e.g. @google2020; @shibayama2021impact]. 
Although there is a broad agreement that these changes 
during the pandemic have played a pivotal role in stopping the spread of the virus 
[@jia2020population; @yang2020taking; @mu2020interplay],
we still have no evidence to what extend these increased levels of working from 
home, and the consequent decrease in commuting flows and the structure of cities, 
will remain post-pandemic.

<!-- Covid-19 and working from home, cities, urban structure -->
Multi-layered digital divides may also intersect with material divides
and the economic geography of the UK.
The regional economic resilience literature underlines the differentiated capacity 
of cities and regions to escape or recover from economic crises 
[@martin2012regional; @kitsos2018economic].
As different places have different industrial and occupational profiles, 
these affect their aggregated potential capacity for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers from some of the highest levels of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
All three levels of digital divides are associated with or shaped by the
geography of the UK. Yet this is the first time that the intersection of digital 
and material divides is relevant to understanding the economic resilience of places 
and large swathes of the population, as digital technologies became an essential
tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity to understand
how internet infrastructure quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive 
video conferencing required in order to avoid the face-to-face contact that could 
increase the spread of infection.
We seek to answer how internet service resilience could contributes to or reduce 
economic resilience when the latter is dependent upon the capability to 
work from home.
We also aim to improve our understanding of the impact of first level digital
divisions on telecommuting, and whether this results in more fundamental third
level digital divisions than has previously been perceived.

@@

In summary, the regression results indicate the complexity of measuring second and 
third level digital divides within spatial aggregations where the geography of 
first level digital divides has been captured using time-series clustering of
experienced broadband upload speeds as a product of reliability not just
availability. Whilst the differences in mean speeds between clusters were not large,
the temporal clustering of internet resilience showed much greater variation, 
and was not spatially dependent upon distance to large urban areas, or relative
location. Internet resilience supports a wide range of small and large urban 
economies, but also fails, or at least frustrates a wide range of other urban 
economies, as demonstrated by the number of tests per capita run in cluster $6$.
Yet our analysis of the likelihood of furlough within the population, especially in 
the smaller suburban clusters $7$, $3$, $5$, and also $8$ shows that digital divides 
were as likely to diverge as intersect during the pandemic, and did not necessarily 
overlap with prior economic divides. With a much greater share of the population 
likely to continue to work from home in the future, and with changing attitudes 
towards residential locations, our analysis suggests that first level digital divides
should be seen as a function of reliability as well as availability, of upload 
speeds as well as download speeds, and of greater spatial variation than might
have previously been considered.

# Discussion and Conclusions {#sec:5}

This paper offers a new perspective on telecommuting from the viewpoint of the
complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods,
a family of unsupervised machine learning techniques which are rarely utilised
in geographical research. Fast, reliable internet connections are
necessary for the population to be able to work from home.
Although not every place hosts individuals in occupations which allow for
telecommuting nor with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that places may depend upon good internet 
reliability as well as connectivity to achieve economic resilience in a period like 
the current pandemic when internet resilience is so vital.

Our analysis demonstrated that the temporal profiles of seven of our
nine clusters had slower upload speeds in the morning than in the evening.
The opposite is likely to have been the norm prior to the pandemic, as level of
demand and bandwidth management is the most common cause of temporal variation in
experienced speeds, and why evening download speeds, rather than daytime upload
speeds, have been used to benchmark the performance of internet services.
Thus, the new patterns can be taken as evidence of widespread telecommuting
and other daytime internet use which changed the temporal profile of
internet activity throughout the UK, not just in areas with more digital industry
or better skills. Furthermore, upload speeds have not previously been seen as 
integral to universal service, considering there has never before been such extreme 
demand for telecommuting and operations such as video calls. Yet those in the 
largest cluster, $6$, clearly experienced lower upload speeds. 
Their speeds were also less reliable, perhaps because they had higher levels of
demand than the second largest cluster, $9$, where many more employees were
furloughed.

Almost all of the largest urban areas in the UK, as well as many smaller cities, 
were split between the largest two clusters -- $6$ and $9$ -- and both contain LADs 
that are centres of the knowledge economy, from Cambridge in the former to Oxford 
in the latter, or have high concentrations of digital businesses, like Reading in 
the former and Milton Keynes in the latter [@technation2017].
Yet there are both intersections and divergences of digital and economic divides.
LADs in cluster $9$ were able to benefit both from reliable internet connections and
populations which were familiar with working from home and could capitalise on their 
digital infrastructure, although overall, they appeared to have a lower rate of
return in the pandemic, with far greater numbers furloughed. 
Is this an indication of both historic economic division, as well as the third level 
digital divide in cities in the North and Midlands unable to capitalise on their 
digital infrastructure?
Meanwhile, did the greater presence in London and the South of LADs in cluster $6$,
as well as the inclusion of many rural areas less impacted by the pandemic help
those in this cluster be more resilient and gain greater returns, despite using 
less reliable internet services?
The nuanced picture we gained through our analysis of the UK case study suggests
that in cluster $1$ as well as $6$, being on the right side of the second level
digital divide had a greater impact on economic resilience and therefore the third
level digital divide, than having quality internet connectivity.

<!-- WFH broader consequences-->
Digital infrastructure which considers upload speeds and working day reliability as 
well as availability are likely to be particularly important in a future where
telecommuting might be a more common means of accessing work and broadband services 
must be fit for purpose, although the long-term effects of such drastic changes in 
telecommuting and attitudes towards working from home are difficult to predict. 
Nevertheless, they span various aspects of economy and society: from changes to
transportation planning due to altered commuting patterns to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713; @ELLDER2020102777], and from productivity
and innovation changes to changes in agglomeration externalities and the
attraction of large cities [@econobs].
Further research may be able to measure the economic resilience of the different
clusters of places discussed in this paper once this pandemic is firmly past.
However, our analysis demonstrates that the economic resilience made possible by
working from home cannot be understood without considering the underpinning digital
divides and cannot be achieved without planning for how the levels of digital,
social and economic divides might intersect.

Early research on Covid-19 and cities speculated the changes that potential extensive
post-Covid working from home patterns might generate for spatial structure: from 
more walkable cities to more localised production and consumption patterns and from
more extensive urban sprawl to decline in public transportation and increased private 
car usage [@batty2020editorial]. Despite the essential role it played during the 
pandemic, less effort has been spent in understanding the current and future role
of digital infrastructure. Our results indicate that probably for the first time 
the future of cities and spatial structure are so much intertwined with digital 
infrastructure. If the post-covid world is a world with extensive working from home, 
then we need to build, among other things, resilient digital infrastructure capable 
of bridging the first layer of the digital divide. Contrary to previous broadband 
business and deployment plans, emphasis should be placed not only on download,
but also on upload speeds [@brake2020lessons. In essence, Covid-19 accentuated the 
old, but more valid than ever argument: digital infrastructure, just like any other 
network infrastructure, only becomes visible when it stops working [@star1999; @tranos2013geography].

>>>>

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds
during the spring 2020 lockdown,
may reinforce or redress the spatial and social dimensions of digital divisions.
We employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
As the quality of internet infrastructure and services, as well as the concentration
of different occupations are spatially dependent and clustered in space,
our approach offers a framework for understanding the impact of and interactions
between digital divisions geographically and socioeconomically.
Our analysis enables us to better understand how the spatial and social distribution
of occupation and online accessibility intersect to enable or hinder 
the practice of telecommuting at a time of extreme demand.
We also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era.

@@

Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
In some aspects, these returns may have increased during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies. Rural areas, predictably, fare worse.
This finding broadly applies in the UK, where studies that also analysed broadband 
speed checks concluded that average speeds are lower in rural areas, something that
has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
In contrast, @riddlesden2014broadband found that levels of deprivation did not correlate with 
first level digital divides.

However, as @blank2018local highlight, variation in individual internet 
availability and uptake is a product of more complex spatial and 
demographic characteristics than simple rurality or urbanisation.
Dense urban areas were shown to suffer more from slowdown during peak hours, 
although these services were more likely to be improved by increased competition 
between providers, such as between new entrants and Virgin Media cable connections
[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
available to only $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider. 
Whether the variation in infrastructure quality and reliability affects the 
spatial footprint of telecommuting has not previously been investigated, in part 
because telecommuting has not previously been a cause of concentrated demand and 
pressure on internet services.

Other intersections between digital divides have been subject to study.
There are indications that those who purchase high speed connections consume
more data of all sorts and use their connections for a greater variety of purposes
[@hauge2011consumer].
There is also a correlation between access to internet services and a
reduction in household transport spend [@bris2017ict].
These studies suggest that better internet services enable households to make 
savings and efficiencies, an example of the first level digital divide reinforcing 
the third level. Such digital intersections and efficiencies have also changed 
transportation, connectivity and the structure of cities during the pandemic.
We now have enough data to confirm the drastic alteration observed in the space-time 
geography of cities around the world in the last year [e.g. @google2020; @shibayama2021impact]. 
Yet, although there is a broad agreement that these changes 
during the pandemic have played a pivotal role in stopping the spread of the virus 
[@jia2020population; @yang2020taking; @mu2020interplay], the extent to which the 
increased levels of working from home, the consequent decrease in commuting flows, 
and the altered structure of cities will remain post-pandemic is the subject
of considerable debate in the literature.

<!-- Covid-19 and working from home, cities, urban structure -->
Meanwhile, multi-layered digital divides may also intersect with material divides
and the economic geography of the UK.
The regional economic resilience literature underlines the differentiated capacity 
of cities and regions to escape or recover from economic crises 
[@martin2012regional; @kitsos2018economic].
As different places have different industrial and occupational profiles, 
these affect their aggregated potential capacity for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south regions of England in terms of skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers from some of the highest levels of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
All three levels of digital divides are associated with or shaped by the
geography of the UK. Yet this is the first time that the intersection of digital 
and material divides is relevant to understanding the economic resilience of places 
and large swathes of the population, as digital technologies became an essential
tool of productivity during the Covid-19 pandemic.

The extreme demand during the pandemic thus provides a new opportunity to understand
how internet infrastructure quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive 
video conferencing required in order to avoid the face-to-face contact that could 
increase the spread of infection.
We seek to answer how internet service resilience could contribute to or reduce 
economic resilience when the latter is dependent upon the capability to 
work from home.
We also aim to improve our understanding of the impact of first level digital
divisions on telecommuting, and whether this results in more fundamental third
level digital divisions than has previously been perceived.

@@

In summary, the regression results indicate the complexity of measuring second and 
third level digital divides within spatial aggregations where the geography of 
first level digital divides has been captured using time-series clustering of
experienced broadband upload speeds as a product of reliability not just
availability. Whilst the differences in mean speeds between clusters were not large,
the temporal clustering of internet resilience showed much greater variation, 
and was not spatially dependent upon distance to large urban areas, or relative
location. Internet resilience supports a wide range of small and large urban 
economies, but also fails, or at least frustrates a wide range of other urban 
economies, as demonstrated by the number of morning peak tests per capita run in 
cluster $6$.
Yet our analysis of the likelihood of furlough within the population, especially in 
the smaller suburban clusters $7$, $3$, $5$, and $8$ shows that digital divides 
were as likely to diverge as intersect during the pandemic, and did not necessarily 
overlap with prior economic divides. With a much greater share of the population 
likely to continue to work from home in the future, and with changing attitudes 
towards residential locations, our analysis suggests that first level digital divides
should be seen as a function of reliability as well as availability, of upload 
speeds as well as download speeds, and of greater spatial variation than might
have previously been considered.

# Discussion and Conclusions {#sec:5}
Our analysis demonstrated that the temporal profiles of seven of our
nine clusters had slower upload speeds in the morning than in the evening.
The opposite is likely to have been the norm prior to the pandemic, as level of
demand and bandwidth management is the most common cause of temporal variation in
experienced speeds, and why evening download speeds, rather than daytime upload
speeds, have been used to benchmark the performance of internet services.
Thus, the new patterns can be taken as evidence of widespread telecommuting
and other daytime internet use which changed the temporal profile of
internet activity throughout the UK, not just in areas with more digital industry
or better skills. Furthermore, upload speeds have not previously been seen as 
integral to universal service, considering there has never before been such extreme 
demand for telecommuting and operations such as video calls. Yet those in the 
largest cluster, $6$, clearly experienced lower upload speeds. 
Their speeds were also less reliable, perhaps because they had higher levels of
demand than the second largest cluster, $9$, where many more employees were
furloughed. Indeed, those with the skills and jobs to work from home were
often left with the least robust services and greatest slowdown.

<!-- WFH broader consequences-->
Therefore, home-based digital infrastructure which considers upload speeds and 
working day reliability as well as availability are likely to be particularly 
important in a future where telecommuting might be a more common means of accessing 
work and broadband services must be fit for purpose. 
Although the long-term effects of such drastic changes in telecommuting and 
attitudes towards working from home are difficult to predict, the reliability of 
home broadband services deserve more consideration than in the past. 
This would represent a switch from previous broadband policies in the UK which 
tended to be aimed at supporting small and medium enterprises on industrial estates 
and business parks []. Such policies also made some assumptions about the quality of 
service in residential markets that could be assured without government intervention 
- an assumption which our analysis suggest is not entirely accurate.

Policy and development proposals should also consider how the potential changes 
wrought by the pandemic span various aspects of economy and society. 
Changes to transportation planning due to altered commuting patterns also imply 
changes in land use and urban planning to accommodate 
people who work from home [@BUDNITZ2020102713; @ELLDER2020102777].
Productivity and innovation changes will reflect upon changes in agglomeration 
externalities and the attraction of large cities [@econobs].
Further research may be able to measure the economic resilience of the different
clusters of places discussed in this paper once this pandemic is firmly past.
However, our analysis demonstrates that the economic resilience made possible by
working from home cannot be understood without considering the underpinning digital
divides and cannot be achieved without planning for how the levels of digital,
social and economic divides might intersect. For example, our results suggest that 
broadband policies cannot improve the economic resilience of places where the 
industrial structure does not align with occupations that incorporate the digital 
skills and capabilities to work from home. Such places instead experienced higher 
proportions of their labour force being placed on furlough.

Early research on Covid-19 and cities also speculated upon the changes that 
potential extensive post-Covid working from home patterns might generate for 
spatial structure: from more walkable cities and more localised production and 
consumption patterns to more extensive urban sprawl, the decline in public 
transportation and increased private car usage [@batty2020editorial]. 
Despite the essential role it played during the pandemic, less effort has been spent 
in understanding the current and future role of digital infrastructure. 
Our results indicate that probably for the first time the future of cities and 
spatial structure are so intertwined with digital infrastructure. 
If the post-covid world is a world with extensive working from home, 
then we need to build, among other things, resilient digital infrastructure capable 
of bridging the first layer of the digital divide. Contrary to previous broadband 
business and deployment plans, emphasis should be placed not only on download,
but also on upload speeds [@brake2020lessons]. In essence, Covid-19 accentuated the 
old argument, now more valid than ever: digital infrastructure, just like any other 
network infrastructure, only becomes visible when it stops working [@star1999; @tranos2013geography].

On the other hand, the nuanced picture we gained through our analysis of the UK 
case study suggests that being on the right side of the second level digital divide 
had a greater impact on economic resilience and therefore the third level digital 
divide, than having quality internet connectivity. 
Our regressions results for cluster $1$ and $6$ are a demonstration of this. 
Our analysis shows both the intersections and divergences of digital and economic 
division.
Almost all of the largest urban areas in the UK, as well as many smaller cities, 
were split between the largest two clusters -- $6$ and $9$ -- and both contain LADs 
that are centres of the knowledge economy, from Cambridge in the former to Oxford 
in the latter, or have high concentrations of digital businesses, like Reading in 
the former and Milton Keynes in the latter [@technation2017].
Yet whilst LADs in cluster $9$ were able to benefit both from reliable internet 
connections and populations familiar with working from home to capitalise on  their 
digital infrastructure, they appeared to have a lower rate of return in 
the pandemic, with far greater numbers furloughed. 
Is this an indication of historic economic division, as well as the third level 
digital divide, in cities in the North and Midlands unable to capitalise on their 
digital infrastructure?
Meanwhile, did the greater presence in London and the South of LADs in cluster $6$,
as well as the inclusion of many rural areas less impacted by the pandemic help
those in this cluster be more resilient and gain greater returns, despite using 
less reliable internet services? Further research would be necessary to prise apart
the detail, particularly in the largest clusters where averages are less distinct.

In conclusion, this paper offers a new perspective on telecommuting from the 
viewpoint of the complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods,
a family of unsupervised machine learning techniques which are rarely utilised
in geographical research. Fast, reliable internet connections are
necessary for the population to be able to work from home.
Although not every place hosts individuals in occupations which allow for
telecommuting nor with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that places may depend upon good internet 
reliability as well as connectivity to achieve economic resilience in a period like 
the current pandemic when internet resilience is so vital.
